How and Why to run a Hack Week

### 2017-12-08 supplementary_material.tex
Pass through some parts of the supplementary materials.

\subsubsection{Obtaining permission for tracking and studying participants}
A goal of most of our previous events was to study the development of a hack week, to observe participant behavior, and quantify the achievement of learning outcomes. To this end, we invited social scientists, interested in mechanisms of scientific collaboration and educational design process, to observe our activities. These scientists then reviewed their observations and and gave valuable feedback about the hack week outcomes. In order to be able to track and study participants, approval from Internal Review Boards is required for workshops organized in the United States. This should be done as early as possible, and having assistance from researchers in the social sciences is invaluable, in particular for hack weeks in fields that generally don't involve human subject research. Ideally, participants should consent to be contacted at the application stage, which opens up the potential of tracking both participants and non-participants and allows for evaluation of the hack week.

@@

Unlike traditional conferences, where the programme is primarily driven by speaker selection, careful goal planning is essential for a successful hack week. The stated goals will continue to determine almost all of the remaining organizational questions, including space, participant selection and workshop evaluation. Is it aimed more at students or mature researchers, or a mix of both? Should the taught component dominate over project work or should it be only a minor programme point? Is the community to be invited more software-focused or should outcomes skew towards publishable results? Does the workshop aim to maximize its output by carrying best practices into universities and institutions that may be particularly underrepresented? These questions, among others, should have a very clear answer before any further workshop organization takes place.

\subsubsection{Hack Week Location}
When considering space, both the broader location (city and country) as well as the specifics (the room in which the workshop will occur) should receive equal consideration. The non-traditional nature of hack weeks means that some students report challenges in receiving their superior's approval to attend. Thus, keeping the costs of attendance low (including travel costs, room and board, and conference fees) should be the highest priority when deciding on a location. Conference centers like the Lorentz Center at the University of Leiden provide funding along with the collaborative space, but have long lead times when applying for their spaces (typically 12-18 months). The same can be true for collaborative spaces within universities and active learning environments, which tend to be popular. Therefore, the search for a location should commence no later than twelve months before the workshop, ideally much earlier.
The space itself should be large and ideally configurable, with moveable tables and chairs, several screens or projectors, and an ample number of white-/chalkboards. A useful guiding principle is to book a space that fits a group at least 25\% larger than that envisioned for the workshop, and the less configurable the space, the larger the extra room should be.

@@

Event size (i.e. the number of attendees) and event durations are important to consider during the design and planning phase of a hack week. All hack weeks held so far have been five days, with group sizes between 40-60 participants. The number of attendees resulted from a number of different design considerations. One practical consideration was the size of the available location: especially the project-focused parts of a hack week generally require a reconfigurable space with more seats than attendees, such that spontaneous organization in sub-groups is easy. However, group size is also important in the context of community building, group dynamics and programme decisions. We limited group size to a maximum of 70 participants, because previous anecdotal evidence has shown us that this is about the maximum group size where interactions between all participants is still possible. We wanted to ensure that all participants had at least met all other participants in order to facilitate community building and keep the group from fracturing into cliques. Another important consideration is representation along the many axes of diversity we considered. For example, a single undergraduate participant might feel like an outsider; admitting three or four can mitigate alienation effects. Similarly, in order to foster active collaboration among many different subfields, we ensured admitting participants with many different types of expertise, some of which overlapped. Having a set of participants that is too small would interfere both with these goals of representation and of fostering interdisciplinary collaborations and networking, thus we would advise admitting at least $\sim$30 or so participants if these objectives are an important part of the hack week.

@@

The hack weeks presented here used different approaches, that nevertheless several common themes and ideas. In particular, they focus on making the selection procedure as quantitative as possible and use the data collected during each stage of the process to critically evaluate the efficacy of the process and its subject to human bias. 
Secondly, much energy is expended early on in the design of the application form, which often use a combination of open-answer and multiple-choice questions to assess candidates' suitability, skill levels and backgrounds. Care must be taken when designing questions assessing skill levels, because options such as ``beginner'' or ``expert'' are have no absolute scaling and may have different meanings to different applicants. They may also be subject to intrinsic biases, if candidates of certain backgrounds systematically underestimate or overestimate their skill levels compared to other candidates. It is thus advisable to calibrate skill levels on concrete milestones. For example, instead of asking participants whether they are ``beginners'' or ``experts'' at machine learning, we instead asked them whether they had done machine learning for class work, had used techniques from machine learning in their own research, or had developed and implemented their own methods. Responding to questions about demographic characteristics should generally be optional, and organizers should consider how these questions might interact adversely with workshop goals, e.g.\ via stereotype threat. For example, Astro Hack Week only asked whether an applicant considered themselves a minority with respect to gender identity within astronomy, rather than asking them to explicitly self-identify within a specific category.

The overall selection procedure for Astro Hack Week comprised a two-step process. In the initial step, we performed a blind assessment for suitability using open-ended questions querying candidates' motivations for attending the workshop, and their goals for attending. We only excluded a small fraction ($<10\%$) of applicants at this step, because our goals aligned with admitting a wide range of different skill levels and backgrounds. Applicants were generally either excluded for being an obvious mismatch with the workshop goals (as in the case of an Android developer applying to work on an unrelated game) or for being duplicate entries. We subsequently used a range of categories queried with multiple-choice questions  on the application to maximize diversity across these categories, which included career stage, gender identity, racial/ethnic identity, geographic origin, programming background, machine learning skills, statistics knowledge, and previous attendance at Astro Hack Week. Targets within each categories were set according to the different workshop goals. For example, we set ambitious goals of admitting 50\% attendees from underrepresented minorities with respect to gender, race and ethnicity with the goal of increasing representation of these groups with astronomy and data science. Similarly, we set targets for other categories (e.g. skill levels) in order to maximize goals around collaboration, networking and fostering interdisciplinary exchange of knowledge. Subsequently, we used a greedy optimization algorithm to jointly optimize over all categories simultaneously. We evaluated our results by directly comparing the characteristics of the selected set of participants with both our targets and the distribution of the set of all candidates. For demographic characteristics specifically, 42\% of participants in both 2016 and 2017 considered themselves a minority with respect to gender identity in astronomy, and 32\% of participants in 2016 (2017: 33\%) considered themselves a minority with respect to racial or ethnic identity. About half of the attendees were students ($\sim$40\% graduate students, $\sim$10\% undergraduate students), another $25\5$ postdoctoral fellow, and the remainder about equally split between research staff, tenured and tenure-track faculty, and participants in non-academic careers.

@@

We have worked to minimize the cost of registration for our events in order to enable broad participation, especially for participants from institutions, countries and demographics that are generally underrepresented at other meetings. Locations that offer some funding or provide free space are preferable to those that do not.
At Astro Hack Week, we have found that travel grants are of crucial importance especially to undergraduates, students from minority backgrounds and participants from underrepresented countries. We found that minority participants disproportionally decline to attend, and often cite lack of sufficient travel funding as the primary reason. In particular if increasing diversity in the more technical aspects of a field is one of the workshop's goals, providing enough travel funding for between five to ten participants can make a significant difference in allowing minority researchers to attend.

@@

At Astro Hack Week, we aim to influence group dynamics in specific ways to allow timid participants to speak up. As mentioned above, we ask a few participants known to us to occasionally ask questions during tutorials, in order to encourage other participants to do the same. We also designate at least two members of the organizing committee whose role it is to facilitate both tutorials and hack sessions. It is their task to for example manage dominant personalities, which can be both a boon and a hindrance to the success of a hack week. Often, these attendees can be effective at generating excitement and engagement, but can also intimidate and dominate discussions. It is up to the organizers to balance the participation of these attendees with the needs of the rest of the group. Similarly, facilitators at Astro Hack Week spend the majority of hacking time circulating the room, connecting participants without a group to a project, help connect groups who run into a problem with experts who can help out, and make sure group work proceeds in such a way that all members feel valued.

>>>>

\subsubsection{Obtaining permission for tracking and studying participants}
A secondary goal of most of our previous events was to study the development of a hack week, to observe participant behavior, and quantify the achievement of learning outcomes. To this end, we invited social scientists, interested in mechanisms of scientific collaboration and educational design process, to observe our activities. These scientists then reviewed their observations and and gave valuable feedback about the hack week outcomes. In order to be able to track and study participants, approval from an Internal Review Board is required. This should be done as early as possible, and having assistance from researchers in the social sciences is invaluable, in particular for hack weeks in fields that generally don't involve human subject research. Ideally, participants should consent to be contacted at the application stage, which opens up the potential of tracking both participants and non-participants and allows for evaluation of the hack week.

@@

Unlike traditional conferences, where the programme is primarily driven by speaker selection, careful goal planning is essential for a successful hack week. The stated goals will continue to determine almost all of the remaining organizational questions, including space, participant selection and workshop evaluation. Is the target audience primarily students, or mature researchers, or a mix of both? Should the taught component dominate over project work or should it be only a minor programme point? Is the community to be invited more software-focused or should outcomes skew towards publishable results? Does the workshop aim to maximize its output by carrying best practices into universities and institutions that may be particularly underrepresented? These questions, among others, should have a very clear answer before any further workshop organization takes place.

\subsubsection{Hack Week Location}
When considering space, both the broader location (city and country) as well as the specifics (the room in which the workshop will occur) should receive equal consideration. The non-traditional nature of hack weeks means that some students report challenges in receiving support from their supervisor or department to attend. Thus, keeping the costs of attendance low (including travel costs, room and board, and conference fees) should be a high priority when deciding on a location. Conference centers like the Lorentz Center at the University of Leiden provide funding along with the collaborative space, but have long lead times when applying for their spaces (typically 12-18 months). The same can be true for collaborative spaces within universities and active learning environments, which tend to be popular. Therefore, the search for a location should commence no later than twelve months before the workshop, ideally much earlier.
The space itself should be large and ideally configurable, with moveable tables and chairs, several screens or projectors, and an ample number of white-/chalkboards. A useful guiding principle is to book a space that fits a group at least 25\% larger than that envisioned for the workshop, and the less configurable the space, the larger the extra room should be.
There should be plenty of power outlets for participants to plug in their computers.
The space should have ready access to wifi, for activities that require an internet connection (e.g., collaborative software development), and organizers should arrange in advance to provide access to visitors from other institutions.
Accessibility should be taken into consideration: is the space accessible for participants who use wheel-chairs? Are accessible restrooms available.

@@

Event size (i.e. the number of attendees) and event durations are important to consider during the design and planning phase of a hack week. All hack weeks held so far have been five days, with group sizes between 40-60 participants. The number of attendees resulted from a number of different design considerations. One practical consideration was the size of the available location: especially the project-focused parts of a hack week generally require a reconfigurable space with more seats than attendees, such that spontaneous organization in sub-groups is easy. However, group size is also important in the context of community building, group dynamics and programme decisions. We limited group size to a maximum of 70 participants, because previous anecdotal evidence has shown us that this is about the maximum group size where interactions between all participants is still possible. We wanted to ensure that all participants had an opportunity to at least meet all other participants, in order to facilitate community building and keep the group from fracturing into cliques. Another important consideration is representation along the many axes of diversity we considered. For example, a single undergraduate participant might feel like an outsider; admitting three or four can mitigate alienation effects. Similarly, in order to foster active collaboration among many different subfields, we ensured admitting participants with many different types of expertise, some of which overlapped. Having a set of participants that is too small would interfere both with these goals of representation and of fostering interdisciplinary collaborations and networking, thus we would advise admitting at least $\sim$30 or so participants if these objectives are an important part of the hack week.

@@

The hack weeks presented here used different approaches, that nevertheless several common themes and ideas. In particular, they focus on making the selection procedure as quantitative as possible and use the data collected during each stage of the process to critically evaluate the efficacy of the process and its subject to human bias.
Secondly, much energy is expended early on in the design of the application form, which often use a combination of open-answer and multiple-choice questions to assess candidates' suitability, skill levels and backgrounds. Care must be taken when designing questions assessing skill levels, because options such as ``beginner'' or ``expert'' are have no absolute scaling and may have different meanings to different applicants. They may also be subject to intrinsic biases, if candidates of certain backgrounds systematically underestimate or overestimate their skill levels compared to other candidates (the Dunning Kruger effect). It is thus advisable to calibrate skill levels on concrete milestones. For example, instead of asking participants whether they are ``beginners'' or ``experts'' at machine learning, we instead asked them whether they had done machine learning for class work, had used techniques from machine learning in their own research, or had developed and implemented their own methods. Similarly, questions about programming skills may ask whether participants can perform a particular task in one of a list of languages  (e.g., "write a module or a library"). Responding to questions about demographic characteristics should generally be optional, and organizers should consider how these questions might interact adversely with workshop goals, e.g.\ via stereotype threat. For example, Astro Hack Week only asked whether an applicant considered themselves a minority with respect to gender identity within astronomy, rather than asking them to explicitly self-identify within a specific category. Meanwhile, neuro hack week opted to ask about gender ethnicity and race, using questions that are standardly used for example, when collecting statistics about study participants.

The overall selection procedure for Astro Hack Week comprised a two-step process. In the initial step, we performed a blind assessment for suitability using open-ended questions querying candidates' motivations for attending the workshop, and their goals for attending. We only excluded a small fraction ($<10\%$) of applicants at this step, because our goals aligned with admitting a wide range of different skill levels and backgrounds. Applicants were generally either excluded for being an obvious mismatch with the workshop goals (as in the case of an Android developer applying to work on an unrelated game) or for being duplicate entries. We subsequently used a range of categories queried with multiple-choice questions  on the application to maximize diversity across these categories, which included career stage, gender identity, racial/ethnic identity, geographic origin, programming background, machine learning skills, statistics knowledge, and previous attendance at Astro Hack Week. Targets within each categories were set according to the different workshop goals. For example, we set ambitious goals of admitting 50\% attendees from underrepresented minorities with respect to gender, race and ethnicity with the goal of increasing representation of these groups with astronomy and data science. Similarly, we set targets for other categories (e.g. skill levels) in order to maximize goals around collaboration, networking and fostering interdisciplinary exchange of knowledge. Subsequently, we used a greedy optimization algorithm to jointly optimize over all categories simultaneously. We evaluated our results by directly comparing the characteristics of the selected set of participants with both our targets and the distribution of the set of all candidates. For demographic characteristics specifically, 42\% of participants in both 2016 and 2017 considered themselves a minority with respect to gender identity in astronomy, and 32\% of participants in 2016 (2017: 33\%) considered themselves a minority with respect to racial or ethnic identity. About half of the attendees were students ($\sim$40\% graduate students, $\sim$10\% undergraduate students), another $25\%$ postdoctoral fellow, and the remainder about equally split between research staff, tenured and tenure-track faculty, and participants in non-academic careers.

Neuro hack week took a different approach: participants provided CV as part of their application, in addition to answering questions about their technical and research background, and writing a short statement of the purpose of their participation. Each application was reviewed by several instructors using 1-10 scales on several criteria: technical and computing experience, research experience, the quality of the statement, an assessment of the contribution of participation to the applicant's career, and an assessment of the degree to which participation would increase the representation of groups that are usually under-represented in neuroscience and computing. The scores were combined, applicants were ranked according to the combined score, and the 40 highest-scoring applicants were then selected to participate. As an example,  this procedure resulted in gender parity in 2016 and near-parity (17 women and 23 men) in 2017. In neuro hack week as well about half of the participants were students (though no undergraduate students participated), about $25\%$ were post-docs, with the remainder including research staff and faculty members.

@@

We have worked to minimize the cost of registration for our events in order to enable broad participation, especially for participants from institutions, countries and demographics that are generally underrepresented at other meetings. Locations that offer some funding or provide free space are preferable to those that do not.
We have found that travel grants are of crucial importance especially to undergraduates, students from minority backgrounds and participants from underrepresented countries. We found that minority participants disproportionally decline to attend, and often cite lack of sufficient travel funding as the primary reason. In particular if increasing diversity in the more technical aspects of a field is one of the workshop's goals, providing enough travel funding for between five to ten participants can make a significant difference in allowing minority researchers to attend.
Funding for some activities can be obtained through commercial vendors.
For example, cloud computing providers have educational programs that provide cloud computing credits for students.
Other corporate sponsorships of certain activities or of the event as a whole can be used to lower participation costs, and make participation more feasible for many participants.
Federal funding agencies are also sometimes able to fund hack weeks and similar activities under traditional funding mechanisms that support training. For example, Neuro hack week 2017 was supported through a grant from the National Institute of Mental Health through their standard R25 training grant mechanism, and this grant continues to support it in extended format in coming years.

@@

From our experience, there are ways in which organizers can influence group dynamics in specific ways to allow timid participants to speak up. As mentioned above, we ask a few participants known to us to occasionally ask questions during tutorials, in order to encourage other participants to do the same. We also designate at least two members of the organizing committee whose role it is to facilitate both tutorials and hack sessions. It is their task to for example manage dominant personalities, which can be both a boon and a hindrance to the success of a hack week. Often, these attendees can be effective at generating excitement and engagement, but can also intimidate and dominate discussions. It is up to the organizers to balance the participation of these attendees with the needs of the rest of the group. Similarly, facilitators at Astro Hack Week spend the majority of hacking time circulating the room, connecting participants without a group to a project, help connect groups who run into a problem with experts who can help out, and make sure group work proceeds in such a way that all members feel valued.
